---
title: "To what extent do gun-laws versus other factors explain the differences in firearm mortality between states?"
author: "Jiaxin Zheng"
format: revealjs
editor: visual
---

```{r}
#| label: load-packages
#| warning: false
#| message: false
# Load necessary libraries
library(tidyverse)
library(sf)
library(usmap)
library(ggthemes)
library(viridis)
library(tidyverse)
library(usmap)
library(ggplot2)
library(gridExtra)
```

```{r}
# Load datasets
mortality <- read_csv('https://raw.githubusercontent.com/Jennyjjxxzz/Data-608_Story_3/refs/heads/main/firearm_mortality2022.csv')
laws <- read_csv('https://raw.githubusercontent.com/Jennyjjxxzz/Data-608_Story_3/refs/heads/main/firearm_laws.csv')
breakdown <- read_csv('https://raw.githubusercontent.com/Jennyjjxxzz/Data-608_Story_3/refs/heads/main/mortality_breakdown.csv')
ownership <- read_csv('https://raw.githubusercontent.com/Jennyjjxxzz/Data-608_Story_3/refs/heads/main/gun-ownership-by-state.csv')
burglary <- read_csv('https://raw.githubusercontent.com/Jennyjjxxzz/Data-608_Story_3/refs/heads/main/burglary.csv')
```

```{r}
# Standardize all data columns names to lowercase
mortality <- mortality %>% rename_with(tolower)
laws <- laws %>% rename_with(tolower)
breakdown <- breakdown %>% rename_with(tolower)
ownership <- ownership %>% rename_with(tolower)
burglary <- burglary %>% rename_with(tolower)
```

```{r}
# Mapping state abbreviations to full names
state_abbreviation_to_full <- c(
  "AL"="Alabama", "AK"="Alaska", "AZ"="Arizona", "AR"="Arkansas", "CA"="California",
  "CO"="Colorado", "CT"="Connecticut", "DE"="Delaware", "FL"="Florida", "GA"="Georgia",
  "HI"="Hawaii", "ID"="Idaho", "IL"="Illinois", "IN"="Indiana", "IA"="Iowa",
  "KS"="Kansas", "KY"="Kentucky", "LA"="Louisiana", "ME"="Maine", "MD"="Maryland",
  "MA"="Massachusetts", "MI"="Michigan", "MN"="Minnesota", "MS"="Mississippi",
  "MO"="Missouri", "MT"="Montana", "NE"="Nebraska", "NV"="Nevada", "NH"="New Hampshire",
  "NJ"="New Jersey", "NM"="New Mexico", "NY"="New York", "NC"="North Carolina",
  "ND"="North Dakota", "OH"="Ohio", "OK"="Oklahoma", "OR"="Oregon", "PA"="Pennsylvania",
  "RI"="Rhode Island", "SC"="South Carolina", "SD"="South Dakota", "TN"="Tennessee",
  "TX"="Texas", "UT"="Utah", "VT"="Vermont", "VA"="Virginia", "WA"="Washington",
  "WV"="West Virginia", "WI"="Wisconsin", "WY"="Wyoming", "DC"="District of Columbia", "District of Columbia" = "District of Columbia"
)
```

```{r}
# Convert state abbreviations to full names in mortality and breakdown datasets
mortality$state <- state_abbreviation_to_full[mortality$state]
breakdown$state <- state_abbreviation_to_full[breakdown$state]
```

```{r}
# Select and clean the columns
mortality <- mortality %>% select(state, firearm_mortality_rate = rate, deaths)
laws <- laws %>% select(state, lawtotal)
breakdown <- breakdown %>% select(state, suicide_rate, homicide_rate)
ownership <- ownership %>% select(state, gunownershippercentage)
burglary <- burglary %>% select(state, burglary_100k)
```

```{r}
# Merge datasets by state
merged_data <- mortality %>% 
  left_join(laws, by = "state") %>%
  left_join(breakdown, by = "state") %>%
  left_join(ownership, by = "state") %>%
  left_join(burglary, by = "state")
```

```{r}
# Convert numeric columns to proper type
numeric_cols <- c("firearm_mortality_rate", "lawtotal", "suicide_rate", "homicide_rate", "gunownershippercentage", "burglary_100k")
merged_data[numeric_cols] <- lapply(merged_data[numeric_cols], as.numeric)

```

```{r}
#| echo: false
#| eval: false
glimpse(merged_data)
merged_data %>% filter(if_any(everything(), is.na))

```

```{r}
merged_data <- merged_data %>% 
  filter(state != "District of Columbia")
```

## Map Showing Firearm Mortality Rate by State

\- It clearly shows higher mortality concentrated in Southern states such as Mississippi and Louisiana. And Western states such as Montana and Alaska.

```{r}
library(plotly)

# Nedd state abbreviations for plotly maps
merged_data <- merged_data %>%
  mutate(state_abbr = state.abb[match(state, state.name)])

# Interactive choropleth map
map <- plot_ly(
  data = merged_data,
  locations = ~state_abbr,
  locationmode = 'USA-states',
  z = ~firearm_mortality_rate,
  text = ~paste(state, "<br>Mortality Rate:", firearm_mortality_rate),
  type = 'choropleth',
  colorscale = 'Aggrnyl',
  colorbar = list(title = "Mortality Rate")
) %>% layout(
  title = "2022 Firearm Mortality Rate by State",
  geo = list(scope = 'usa')
)

map
```

## Gun Control Laws Map by State

\- States with very strict laws (like California, New York) typically have lower firearm mortality rates.

```{r}
# Categorize Gun Control clearly based on lawtotal
merged_data <- merged_data %>%
  mutate(gun_control_level = case_when(
    lawtotal >= 80 ~ "Very Strict",
    lawtotal >= 50 ~ "Strict",
    lawtotal >= 20 ~ "Moderate",
    lawtotal >= 10 ~ "Loose",
    TRUE ~ "Very loose"
  ))
merged_data_no_na <- merged_data %>%
  drop_na(gun_control_level)

```

```{r}
map2 <- plot_usmap(data = merged_data, values = "gun_control_level", color = "orange") + 
  scale_fill_manual(name = "Gun Control", 
                    values = c("Very loose"="#F0F0F0", 
                               "Loose"="#C0C0C0",
                               "Moderate"="#808080", 
                               "Strict"="#404040", 
                               "Very Strict"="#000000")) +
  labs(title = "Gun Control Laws by State") +
  theme_minimal()
map2
```

## Firearm Mortality Rate vs. Firearm Laws

\- It shows more firearm laws correlate with lower firearm mortality.(But the trend is weak, that laws alone do not explain the whole picture)

```{r}
# Relationship Between Firearm Laws and Mortality Rate
p1 <- merged_data %>%
  ggplot(aes(x = lawtotal, y = firearm_mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = "Firearm Mortality Rate vs. Number of Firearm Laws", x = "Total Firearm Laws", y = "Firearm Mortality Rate")

p1
```

## Firearm Mortality Rate vs. Gun Ownership Percentage

\- Higher gun ownership clearly linked to higher firearm mortality.

```{r}
# Relationship Between Gun Ownership and Mortality Rate
p2 <- merged_data %>%
  ggplot(aes(x = gunownershippercentage, y = firearm_mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = "Firearm Mortality Rate vs. Gun Ownership Percentage", x = "Gun Ownership (%)", y = "Firearm Mortality Rate")

p2
```

## Gun Ownership Percentage

\- Northeast states like Montana, Wyoming, and Idaho have gun ownership rate up to 66% and they are followed by southern states such as Mississippi and Louisiana have gun ownership rates as high as 55%.

```{r}
merged_data$gunownershippercentage <- as.numeric(merged_data$gunownershippercentage)

map3 <- plot_ly(
  data = merged_data,
  locations = ~state_abbr,
  locationmode = 'USA-states',
  z = ~gunownershippercentage,  # Make sure this is correctly mapped
  text = ~paste(state, "<br>Gun Ownership(%):", gunownershippercentage),
  type = 'choropleth',
  colorscale = 'Aggrnyl',  # Green color scale
  colorbar = list(title = "Gun Ownership Percentage"),
  zmin = min(merged_data$gunownershippercentage, na.rm = TRUE),  # Ensures correct scaling
  zmax = max(merged_data$gunownershippercentage, na.rm = TRUE)
) %>% layout(
  title = "Gun Ownership Percentage by State",
  geo = list(scope = 'usa')
)

map3

```

## Firearm Suicide Rate by State

\- Northwestern states, which have higher gun ownership rates, also tend to have higher suicide rates. These states have more relaxed gun laws.

\- Surprisingly, Mississippi and Louisiana have relatively low suicide rates despite their high gun ownership.

```{r}
merged_data <- merged_data %>%
  mutate(state_abbr = state.abb[match(state, state.name)]) %>%
  drop_na(state_abbr, suicide_rate, homicide_rate)
```

```{r}
map4 <- plot_ly(
  data = merged_data,
  locations = ~state_abbr,
  locationmode = 'USA-states',
  z = ~suicide_rate,
  text = ~paste(state, "<br>Suicide Rate:", suicide_rate),
  type = 'choropleth',
  colorscale = 'Aggrnyl',
  colorbar = list(title = "Suicide Rate per 100k")
) %>% layout(
  title = "Firearm Suicide Rate by State",
  geo = list(scope = 'usa')
)
map4
```

## Firearm Homicide Rate by State

\- Southern states like Mississippi and Louisiana have higher homicide rate, surprisingly they are follow by Illinois, New York, Maryland, and California.

```{r}
map5 <- plot_ly(
  data = merged_data,
  locations = ~state_abbr,
  locationmode = 'USA-states',
  z = ~homicide_rate,
  text = ~paste(state, "<br>Homicide Rate:", homicide_rate),
  type = 'choropleth',
  colorscale = 'Aggrnyl',
  colorbar = list(title = "Homicide Rate per 100k")
) %>% layout(
  title = "Firearm Homicide Rate by State",
  geo = list(scope = 'usa')
)
map5
```

## Conclusion

-   States with stricter gun laws consistently show lower firearm mortality rates. However, gun laws alone don't tell the whole story, higher gun ownership may be the cultural factor.

-   While higher gun ownership is strongly linked to increased firearm mortality, but the relationship varies across different types of gun related deaths. In northwestern states, higher gun ownership is associated with higher suicide rates, whereas in southern states, it is linked to higher homicide rates.
